Bayesian Sparsification Methods for Deep Complex-valued Networks

### 2020-06-01 paper/paper.tex
minor typo fixes

Inherently complex-valued nature of data in signal processing, specifically, acoustic or radio
frequency signal analysis, has been the key driver behind the adoption of $\cplx$-networks.
%
\citet{hirose_complex-valued_2009} argues that $\cplx$-valued networks merit from the
simultaneous phase rotation and amplitude adjustment, intrinsic to $\cplx$ arithmetic,
which reduce ineffective degrees of freedom in comparison with an double-dimensional
$\real$-valued networks. Their study lists numerous applications related to acoustic
and radio signal processing and demonstrates superiority of $\cplx$-networks in the
task of shallow landmine detection using ground penetrating radar imaging, in spite of
inherently higher arithmetic complexity of $\cplx$-networks, stemming from $3-4$ floating
point operations per single $\cplx$ multiplication. More recently $\cplx$-networks have
been applied to music transcription and speech recognition \citep{trabelsi_deep_2017}.
\citet{arjovsky_unitary_2016} propose $\cplx$-valued unitary-RNNs to deal with exploding~/~vanishing
gradients and apply the recurrent models to long-term dependence prediction tasks.
\citet{wisdom_full-capacity_2016} expand on their work by optimizing recurrent weights on
the $\cplx^{n \times n}$ Stiefel manifold. And \citet{wolter_complex_2018} investigate
different $\cplx$-valued gating mechanisms for unitary-RNNs and conduct experiments on
motion prediction and music transcription. \citet{yang_complex_2019} develop complex
transformer, with $\cplx$-attention and complex encoder-decoder, and apply it to music
transcription and wireless signal classification. \citet{tarver_design_2019} achieve
significantly lower out-of-band performance with a $\cplx$-valued network in Digital Signal
Predistortion task.

@@

Save for the relatively simple cases, the posterior distribution is intractable, and
therefore exactness Bayesian Inference is traded for tractability and scalability of
Variational Bayesian Inference. Instead of deriving the posterior $p(\omega \mid D)$,
this approach postulates an optimization problem for finding an approximation $q$,
which is close to $p(\omega \mid D)$ in terms of a proximity score $\rho$
\begin{equation}  \label{eq:variational-progam}
  q_*
    \in \arg \min_{\theta} \rho\bigl(
      q_\theta(\omega), p(\omega \mid D)
    \bigr)
    \,,
\end{equation}
over a tractable parametric family of distributions on the parameter space $
  \mathcal{Q} = \{Q_\theta(d\omega) \colon \theta \in \Theta\}
$, $\theta$ -- generic variational parameter. The variational approximation class
$\mathcal{Q}$ is chosen so that its members have tractable densities $
  q_\theta(\omega) d\omega
$, can be sampled from and have either tractable $\log$-derivatives $
  \omega \mapsto \nabla_\theta \log q_\theta(\omega)
$ or can be represented as differentiable transformations of non-parametric noise
\citep{williams_simple_1992,kingma_auto-encoding_2014,figurnov_implicit_2019}.
% (reparameterization trick, pathwise gradient) $q_{\theta}(d\omega)$ is a push-forward
% of $p(d\varepsilon)$ by a differentiable map $g(\varepsilon; \theta)$.
% % http://stillbreeze.github.io/REINFORCE-vs-Reparameterization-trick/#fn:1
%
The most common choice for $\rho$ is Kullback-Leibler divergence of density $q$ from $p$,
\begin{equation}  \label{eq:kl-div-def}
  KL(q \| p)
    % = \int \frac{dQ}{dP} \log{\frac{dQ}{dP}} P(d\omega)
    % = \int \frac{q(\omega)}{p(\omega)} \log{\frac{q(\omega)}{p(\omega)}} p(\omega) d\omega
    % = \mathbb{E}_{\omega \sim q}
    %   \log\frac{q(\omega)}{p(\omega)}
    = \mathbb{E}_{\omega \sim q}
      \log{q(\omega)}
    - \mathbb{E}_{\omega \sim q}
      \log{p(\omega)}
    \,,
\end{equation}
% and, despite it not being a metric and requiring distributions with nested support,
the main reason being that it satisfies the following identity for well-behaved densities:
% $(q, p, \pi)$
\begin{multline}  \label{eq:kl-div-master}
  % \log p(D)
    % = \mathbb{E}_{q(\omega)} \log \frac{p(D, \omega)}{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{q(\omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
    % - \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{\pi(\omega)}
    % = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
  \log p(D)
    - KL(q \| p(\omega \mid D))
    \\ = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    - KL(q \| \pi)
    \,.
\end{multline}
This leads to an equivalent problem for \eqref{eq:variational-progam} of maximizing
the \textit{evidence lower bound} (ELBO)
\begin{equation}  \label{eq:elbo_general}
  \mathcal{L}(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      + \mathbb{E}_{\omega \sim q_{\theta}}
        \log p_{\phi}(D \mid \omega)
      % \mathbb{E}_{\omega \sim q_{\theta}}
        % \log\frac{q_\theta(\omega)}{\pi_{\lambda}(\omega)}
  % = \mathbb{E}_{\omega \sim q_{\theta}}
  %     \log{p_{\phi}(D \mid \omega) \pi_{\lambda}(\omega)}
  %   + \mathbb{H}(q_{\theta})
  \,,
\end{equation}
where $\lambda$, $\theta$ and $\phi$ are optimization parameters of the prior $\pi$,
approximation $q$ and the likelihood $p(D\mid \omega)$, respectively. Other objectives
are possible, besides \eqref{eq:elbo_general}, provided $p(\omega \mid D)$ is evaluated
only through $\log p(D \mid \omega)$ and $\log \pi(\omega)$, \citep{ranganath_operator_2018}.

@@

During the \textit{``sparsify''} stage we make every layer Bayesian and apply $\cplx$-variational
dropout sec.~\ref{ssub:vd_prior} (or sec.~\ref{ssub:ard_prior}) or its $\real$ counterpart.
We inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:elbo} to study the attainable sparsity levels:
\begin{equation}  \label{eq:elbo_with_coef}
  - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
  + \frac1{M} \sum_{k=1}^M
      \log p_{\phi}(z_{i_k} \mid g(\varepsilon_k; \theta))
  \,.
  \tag{\ref{eq:elbo}'}
\end{equation}
By varying $C$ we adjust the importance of the sparsifying prior in learning $q_\theta(W)$,
meaning that higher value yields more irrelevant parameters during in inference and makes
layers sparser upon termination of this stage.

@@

The overarching pattern in all experiments on MNIST-like datasets is that both ARD and VD yield
compress similarly and ARD offers marginal advantage in terms of performance after fine-tuning.
Fourier features perform marginally worse than raw image data for each model across all
compression rates, and their trade-off for the dense network has a hat-like shape, contrasting
the raw image features, for which the accuracy stays at one level for moderate rates,
and then abruptly decays in high compression regime.

@@

% draw some conclusions
Figure~\ref{fig:figure__cifar10__trade-off} shows that, although ARD method offers slightly
less compression, it makes up for it by marginally better accuracy post fine-tune. VGG16 and
$\cplx$-VGG16 exhibit similar declining compression-accuracy trade-off, but due to higher
capacity $\cplx$-valued network yields higher accuracy \citep{monning_evaluation_2018}. Unlike
MNIST, fine-tuning stage significantly improves the test performance, which, we speculate, might
be due to dataset augmentation and deeper model.

@@

\begin{figure}[!t]
  \centering
  \begin{subfigure}[b]{1.\columnwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=\columnwidth]{figure__fine-tune_fx__early.pdf}
  \end{subfigure}
  \caption{%
    Early stopping epoch at fine-tuning stage.
  }
  \label{fig:musicnet__early_stopping}
\end{figure}

>>>>

Inherently complex-valued nature of data in signal processing, specifically, acoustic or radio
frequency signal analysis, has been the key driver behind the adoption of $\cplx$-networks.
%
\citet{hirose_complex-valued_2009} argues that $\cplx$-valued networks merit from the
simultaneous phase rotation and amplitude adjustment, intrinsic to $\cplx$ arithmetic,
which reduce ineffective degrees of freedom in comparison with double-dimensional
$\real$-valued networks. Their study lists numerous applications related to acoustic
and radio signal processing and demonstrates superiority of $\cplx$-networks in the
task of shallow landmine detection using ground penetrating radar imaging, in spite of
inherently higher arithmetic complexity of $\cplx$-networks, stemming from $3-4$ floating
point operations per single $\cplx$ multiplication. More recently $\cplx$-networks have
been applied to music transcription and speech recognition \citep{trabelsi_deep_2017}.
\citet{arjovsky_unitary_2016} propose $\cplx$-valued unitary-RNNs to deal with exploding~/~vanishing
gradients and apply the recurrent models to long-term dependence prediction tasks.
\citet{wisdom_full-capacity_2016} expand on their work by optimizing recurrent weights on
the $\cplx^{n \times n}$ Stiefel manifold. And \citet{wolter_complex_2018} investigate
different $\cplx$-valued gating mechanisms for unitary-RNNs and conduct experiments on
motion prediction and music transcription. \citet{yang_complex_2019} develop complex
transformer, with $\cplx$-attention and complex encoder-decoder, and apply it to music
transcription and wireless signal classification. \citet{tarver_design_2019} achieve
significantly lower out-of-band performance with a $\cplx$-valued network in Digital Signal
Predistortion task.

@@

Save for the relatively simple cases, the posterior distribution is intractable, and
therefore exactness Bayesian Inference is traded for tractability and scalability of
Variational Bayesian Inference. Instead of deriving the posterior $p(\omega \mid D)$,
this approach postulates an optimization problem for finding an approximation $q$,
which is close to $p(\omega \mid D)$ in terms of a proximity score $\rho$
\begin{equation}  \label{eq:variational-progam}
  q_*
    \in \arg \min_{\theta} \rho\bigl(
      q_\theta(\omega), p(\omega \mid D)
    \bigr)
    \,,
\end{equation}
over a tractable parametric family of distributions on the parameter space $
  \mathcal{Q} = \{Q_\theta(d\omega) \colon \theta \in \Theta\}
$, $\theta$ -- generic variational parameter. The variational approximation class
$\mathcal{Q}$ is chosen so that its members have tractable densities $
  q_\theta(\omega) d\omega
$, can be sampled from and have either tractable $\log$-derivatives $
  \omega \mapsto \nabla_\theta \log q_\theta(\omega)
$ or can be represented as differentiable transformations of non-parametric noise
\citep{williams_simple_1992,kingma_auto-encoding_2014,figurnov_implicit_2019}.
% (reparameterization trick, pathwise gradient) $q_{\theta}(d\omega)$ is a push-forward
% of $p(d\varepsilon)$ by a differentiable map $g(\varepsilon; \theta)$.
% % http://stillbreeze.github.io/REINFORCE-vs-Reparameterization-trick/#fn:1
%
The most common choice for $\rho$ is Kullback-Leibler divergence of density $q$ from $p$,
\begin{equation}  \label{eq:kl-div-def}
  KL(q \| p)
    % = \int \frac{dQ}{dP} \log{\frac{dQ}{dP}} P(d\omega)
    % = \int \frac{q(\omega)}{p(\omega)} \log{\frac{q(\omega)}{p(\omega)}} p(\omega) d\omega
    % = \mathbb{E}_{\omega \sim q}
    %   \log\frac{q(\omega)}{p(\omega)}
    = \mathbb{E}_{\omega \sim q}
      \log{q(\omega)}
    - \mathbb{E}_{\omega \sim q}
      \log{p(\omega \mid D)}
    \,,
\end{equation}
% and, despite it not being a metric and requiring distributions with nested support,
the main reason being that it satisfies the following identity for well-behaved densities:
% $(q, p, \pi)$
\begin{multline}  \label{eq:kl-div-master}
  % \log p(D)
    % = \mathbb{E}_{q(\omega)} \log \frac{p(D, \omega)}{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{q(\omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
    % - \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{\pi(\omega)}
    % = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
  \log p(D)
    - KL(q \| p(\omega \mid D))
    \\ = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    - KL(q \| \pi)
    \,.
\end{multline}
This leads to an equivalent problem for \eqref{eq:variational-progam} of maximizing
the \textit{evidence lower bound} (ELBO)
\begin{equation}  \label{eq:elbo_general}
  \mathcal{L}(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      + \mathbb{E}_{\omega \sim q_{\theta}}
        \log p_{\phi}(D \mid \omega)
      % \mathbb{E}_{\omega \sim q_{\theta}}
        % \log\frac{q_\theta(\omega)}{\pi_{\lambda}(\omega)}
  % = \mathbb{E}_{\omega \sim q_{\theta}}
  %     \log{p_{\phi}(D \mid \omega) \pi_{\lambda}(\omega)}
  %   + \mathbb{H}(q_{\theta})
  \,,
\end{equation}
where $\lambda$, $\theta$ and $\phi$ are optimization parameters of the prior $\pi$,
approximation $q$ and the likelihood $p(D\mid \omega)$, respectively. Other objectives
are possible, besides \eqref{eq:elbo_general}, provided $p(\omega \mid D)$ is evaluated
only through $\log p(D \mid \omega)$ and $\log \pi(\omega)$, \citep{ranganath_operator_2018}.

@@

During the \textit{``sparsify''} stage we make every layer Bayesian and apply $\cplx$-variational
dropout sec.~\ref{ssub:vd_prior} (or sec.~\ref{ssub:ard_prior}) or its $\real$ counterpart.
We inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:elbo} to study the attainable sparsity levels:
\begin{equation}  \label{eq:elbo_with_coef}
  - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
  + \frac1{M} \sum_{k=1}^M
      \log p_{\phi}(z_{i_k} \mid g(\varepsilon_k; \theta))
  \,.
  \tag{\ref{eq:elbo}'}
\end{equation}
By varying $C$ we adjust the importance of the sparsifying prior in learning $q_\theta(W)$,
meaning that higher value yields more irrelevant parameters during inference and makes
layers sparser upon termination of this stage.
% RFC: citation needed fot this "trick". Cite $\beta$-VAE maybe?

@@

The overarching pattern in all experiments on MNIST-like datasets is that both ARD and VD
compress similarly and ARD offers marginal advantage in terms of performance after fine-tuning.
Fourier features perform marginally worse than raw image data for each model across all
compression rates, and their trade-off for the dense network has a hat-like shape, contrasting
the raw image features, for which the accuracy stays at one level for moderate rates,
and then abruptly decays in high compression regime.

@@

% draw some conclusions
Figure~\ref{fig:figure__cifar10__trade-off} shows that, although ARD method offers slightly
less compression, it makes up for it by marginally better accuracy post fine-tune. VGG16 and
$\cplx$-VGG16 exhibit similar declining compression-accuracy trade-off, but due to higher
capacity $\cplx$-valued network yields higher accuracy \citep{monning_evaluation_2018}. Unlike
MNIST, fine-tuning stage significantly improves the test performance, which, we speculate, might
be due to dataset augmentation and deeper model.
% reword this sentence

@@

\begin{figure}[!t]
  \centering
  \begin{subfigure}[b]{1.\columnwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=\columnwidth]{figure__fine-tune_fx__early__compression.pdf}
  \end{subfigure}
  \caption{%
    Early stopping epoch at fine-tuning stage.
  }
  \label{fig:musicnet__early_stopping}
\end{figure}
